shorts: An R Package for Modeling Short Sprints

### 2021-07-21 paper.Rmd
Added section to want about predict_XXX issues with distance correction model when distance correction <>0m

---
title: "shorts: An R Package for Modeling Short Sprints"
preprint: false
date: "`r Sys.Date()`"
author: 
  - name: Mladen Jovanović
    affiliation: 1
    corresponding: true
    email: coach.mladen.jovanovic@gmail.com
  - name: Jason D. Vescovi
    affiliation: 2
    corresponding: false
    email: vescovij@gmail.com
affiliation:
  - code: 1
    address: Faculty of Sport and Physical Education, University of Belgrade, Serbia
  - code: 2
    address: Faculty of Kinesiology and Physical Education, Graduate School of Exercise Science, Toronto, ON Canada
abstract: >
  Short sprint performance is one of the most distinguishable and admired physical traits in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). The most common methods to assess short sprint performance are with a radar gun or timing gates. In this paper, we: 1) provide the **shorts** package that can model sprint timing data from these two sources; 2) discuss potential issues with assessing sprint time (synchronization and flying start, respectively); and 3) provide model definitions within the **shorts** package to help alleviate errors within the subsequent parameter outcomes.
header-includes: >
  \usepackage{lipsum}
  \usepackage[htt]{hyphenat}
  \usepackage{booktabs}
  \usepackage{microtype}
preamble: >
  \usepackage{amsmath}
citation_package: natbib
bibliography: [references.bib, packages.bib]
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
    extra_dependencies: ["flafter"]
  rticles::peerj_article: default
link-citations: yes
editor_options:
  chunk_output_type: console
---

@@

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  fig.retina = 0.8, # figures are either vectors or 300 dpi diagrams
  dpi = 300,
  out.width = "90%",
  fig.align = "center",
  fig.width = 5,
  fig.height = 5 * 0.618, # 1 / phi
  fig.show = "hold",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  width = 65
)

# automatically create a bib database for R packages
# knitr::write_bib(c(
#  .packages(), "bookdown", "knitr", "rmarkdown", "shorts", "nlme", "LambertW"
# ), "packages.bib")

@@

Although TAU is used in the equations, and later estimated, it is preferred to use MAC instead since it is easier to grasp, particularly for less math inclined coaches.

@@

Plotting acceleration against velocity (Figure \@ref(fig:four-athletes-profile)), we will get *Acceleration-Velocity Profile*, which is linear, according to the mathematical model. If the athlete's body mass (kg) is known, as well as additional air resistance parameters (see [Air resistance and the calculation of force and mechanical power] section of this paper), *Force-Velocity Profile* can be estimated (see [Force-Velocity profile] section of this paper). 
```{r four-athletes-profile, echo=FALSE, fig.cap="Acceleration-Velocity profile of four athletes with different MSS and MAC parameters."}
gg <- ggplot(
  kinematics,
  aes(x = velocity, y = acceleration, color = Athlete)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  ylab(expression("Acceleration (ms"^-2 * ")")) +
  xlab(expression("Velocity (ms"^-1 * ")")) +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40 m), (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \@ref(eq:velocity-time) is performed in **shorts** package using non-linear least squares regression implemented in the `nls()` function in the **base R** [@R-base] and `nlme()` function in the **nlme** package [@R-nlme] for the mixed-effect models.

## Estimating short sprint parameters using split times

Let's consider an example of an athlete with MSS equal to 9 $ms^{-1}$, TAU equal to 1.3, and MAC equal to 6.92 $ms^{-2}$ performing 40m sprint with timing gates positioned at each 10m split. For split times, distance is a predictor, and time is the outcome variable, thus the equation \@ref(eq:velocity-time) becomes:

@@

$W$ in equation \@ref(eq:time-distance) represents Lambert's W function [@R-LambertW]. MSS and TAU parameters are estimated using `model_using_splits()` function:

@@

Additional information about mixed-effects model performed using the `nlme` package [@R-nlme] can be obtained using `summary()` function. S3 method `coef()` when applied on mixed-model result will return both the fixed and random effects. To create a simple plot of the model, use S3 `plot()` method.

@@

@@

## Problems at the start when using split times

@@

John represents the *theoretical model*, in which we assume that the initial force production and the timing initiation are perfectly synchronized. Jack, on the other hand, represents a *practical model*, and decides to move slightly behind the initial timing gate (i.e. for 0.5 m) and use body rocking to initiate the sprint start. In other words, Jack is using a *flying start*, a common scenario when testing field sports athletes.

@@

Here is a graphical representation of the sprint splits (please refer to the Supplemental files for the R code):

@@

To explore this further, we have run a simple simulation with 5, 10, 20, 30, 40, and 50 m splits, with MSS and MAC varying from 6 to 9 ($ms^{-1}$ and $ms^{-2}$ respectively), and flying start distance varying from 0 to 1 m (please refer to the Supplemental files for the R code). With a simulation dataset, we can check the model estimates and predictions, given the flying start distance.

```{r include=FALSE}
sim_df <- expand.grid(
  MSS = c(6, 7, 8, 9),
  MAC = c(6, 7, 8, 9),
  flying_start_distance = c(
    seq(0, 0.001, length.out = 20),
    seq(0.001, 0.01, length.out = 20),
    seq(0.01, 0.1, length.out = 20),
    seq(0.1, 1, length.out = 20)
  ),
  distance = c(5, 10, 20, 30, 40, 50)
)

@@

# Prediction residuals
model_df$residuals <- model_df$predicted_time - model_df$time
```

The following figure demonstrates the effect of flying start distance on estimated MSS:

```{r echo=FALSE}
# Estimates plot
df <- model_df %>%
  group_by(MSS, TAU, flying_start_distance) %>%
  slice(1) %>%
  mutate(
    MSS_string = paste("MSS =", MSS),
    TAU_string = paste("TAU =", TAU),
    MAC_string = paste("MAC = ", round(MAC, 2)),
    PMAX_string = paste("PMAX = ", round(PMAX, 2))
  )

# MSS
ggplot(
  df,
  aes(x = flying_start_distance, y = est_MSS, color = MAC_string)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MSS (m/s)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

As can be seen from the figure, MSS is underestimated as flying start distance increases. The following image demonstrates the effect of flying start distance on estimated MAC:

```{r echo=FALSE}
# MAC
ggplot(
  df,
  aes(x = flying_start_distance, y = est_MAC, color = MSS_string)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~MAC_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MAC (m/s/s)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

MAC (and also TAU) are highly affected by the flying start distance, and from the figure we can notice that MAC is overestimated as flying start distance increases.

And finally, the following image demonstrates the effect of flying start distance on estimated PMAX:

```{r echo=FALSE}
# PMAX
ggplot(
  df,
  aes(x = flying_start_distance, y = est_PMAX, color = MSS_string)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~MAC_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated PMAX (W/kg)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

Estimated PMAX is also overestimated as flying start distance increases.

Model residuals are also affected by flying start distance. The shape of residuals distribution depends on number and splits utilized (e.g., 10, 20, 30, 40 m versus 5, 15, 30 m), but here we can see the effect of the flying start distance on the model residuals per split distance utilized in our simulation:

```{r echo=FALSE}
# Residuals
model_df <- model_df %>%
  mutate(
    MSS_string = paste("MSS =", MSS),
    TAU_string = paste("TAU =", TAU),
    MAC_string = paste("MAC = ", round(MAC, 2)),
    PMAX_string = paste("PMAX = ", round(PMAX, 2)),
    group = paste(MSS, MAC, flying_start_distance)
  )

ggplot(
  model_df,
  aes(
    y = residuals,
    x = distance,
    color = flying_start_distance,
    group = group
  )
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  facet_grid(MSS_string ~ MAC_string) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)") +
  theme(legend.position = "top") +
  labs(color = "Flying start distance")
```

If we merge individual facets (i.e., combinations of MSS and MAC), we can get simpler figure conveying issues with residuals when there is a flying start:

```{r echo=FALSE}
ggplot(
  model_df,
  aes(
    y = residuals,
    x = distance,
    color = flying_start_distance,
    group = group
  )
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)") +
  theme(legend.position = "top") +
  labs(color = "Flying start distance")
```

@@

In Jack's case, both +0.3 s fixed time correction and time correction estimation yield parameters closer to John's (i.e. true parameters).

Another model definition, which is a novel approach implemented in the **shorts** package, is to utilize *distance correction*, besides time correction. Thus, equation \@ref(eq:time-distance) becomes:

\begin{equation}
  t(d) = TAU \times W(-e^{\frac{-d + distance \; correction}{MSS \times TAU}} - 1) + \frac{d + distance \; correction}{MSS} + TAU - time \; correction (\#eq:distance-correction)
\end{equation}

@@

```{r}
jack_profile_distance_correction <- model_using_splits_with_corrections(
  distance = split_times$distance,
  time = split_times$jack_time
)

jack_parameters <- rbind(
  coef(john_profile),
  coef(jack_profile),
  coef(jack_profile_fixed_time_short),
  coef(jack_profile_fixed_time_long),
  coef(jack_profile_time_estimated),
  coef(jack_profile_distance_correction)
)

rownames(jack_parameters) <- c(
  "John",
  "Jack - No corrections",
  "Jack - Fixed time correction (+0.3s)",
  "Jack - Fixed time correction (+0.5s)",
  "Jack - Estimated time correction",
  "Jack - Estimated distance correction"
)

@@

As can be seen from the results, adding distance correction results in correctly estimating Jack's sprint parameters. There are a few issues with this model definition. Besides being novel and still not validated with actual data, distance correction model has four parameters to estimate, which implies that at least five sprint splits are needed. This imposes practical limitations, since acquiring six timing gate (one for the start and five for splits) might be practically troublesome. One strategy that is sometimes implemented is adding zeros to the sample (i.e., $t=0$ and $d=0$), which increase the number of observations. Unfortunately, this strategy should not be implemented, as explained later in the [Should we add zero to the sample?] section of this paper.

We will get back to these issues later, but we can examine how these models (no correction, fixed time correction, estimated time correction, and estimated time and distance correction) perform using simulated data with varying flying start distance (please refer to the Supplemental files for the R code).

```{r echo=FALSE}
pred_wrapper <- function(data) {
  no_correction <- model_using_splits(
    distance = data$distance,
    time = data$time
  )

@@

  time_correction <- model_using_splits_with_time_correction(
    distance = data$distance,
    time = data$time,
    control = nls.control(tol = 1)
  )

  time_dist_correction <- model_using_splits_with_corrections(
    distance = data$distance,
    time = data$time,
    control = nls.control(tol = 1)
  )

  params <- rbind(
    data.frame(
      model = "No correction",
      t(coef(no_correction))
    ),
    data.frame(
      model = "Fixed correction +0.3s",
      t(coef(fixed_correction_short))
    ),
    data.frame(
      model = "Fixed correction +0.5s",
      t(coef(fixed_correction_long))
    ),
    data.frame(
      model = "Time correction",
      t(coef(time_correction))
    ),
    data.frame(
      model = "Time and distance correction",
      t(coef(time_dist_correction))
    )
  )

  colnames(params) <- c(
    "model", "est_MSS", "est_TAU", "est_MAC", "est_PMAX",
    "est_time_correction", "est_distance_correction"
  )

  df <- expand_grid(
    data,
    params
  )

  df$predicted_time <- predict_time_at_distance(
    distance = df$distance,
    MSS = df$est_MSS,
    TAU = df$est_TAU,
    time_correction = df$est_time_correction,
    distance_correction = df$est_distance_correction
  )

  df$residuals <- df$predicted_time - df$time
  return(df)
}

# estimated parameters and predicted time
model_df <- sim_df %>%
  group_by(MSS, TAU, flying_start_distance) %>%
  do(pred_wrapper(.)) %>%
  ungroup()
```

As can be seen from the next figure, the estimated time correction model estimates MSS almost perfectly, while the estimated time and distance correction model estimates MSS perfectly.

```{r echo=FALSE}
model_df$model <- factor(
  model_df$model,
  levels = c(
    "No correction",
    "Fixed correction +0.3s",
    "Fixed correction +0.5s",
    "Time correction",
    "Time and distance correction"
  )
)
# Estimates plot
df <- model_df %>%
  group_by(MSS, TAU, flying_start_distance, model) %>%
  slice(1) %>%
  mutate(
    MSS_string = paste("MSS =", MSS),
    TAU_string = paste("TAU =", TAU),
    MAC_string = paste("MAC = ", round(MAC, 2)),
    PMAX_string = paste("PMAX = ", round(PMAX, 2))
  )

# MSS
ggplot(
  df,
  aes(x = flying_start_distance, y = est_MSS, color = model)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_grid(MSS_string ~ MAC_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MSS (m/s)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

Similar outcomes are observed for the MAC parameter. The time and distance corrections model performs perfectly, while the time correction model performs almost as good.

```{r echo=FALSE}
# MAC
ggplot(
  df,
  aes(x = flying_start_distance, y = est_MAC, color = model)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_grid(MAC_string ~ MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MAC (m/s/s)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

PMAX demonstrates the same properties as MSS and MAC.

```{r echo=FALSE}
# PMAX
ggplot(
  df,
  aes(x = flying_start_distance, y = est_PMAX, color = model)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_grid(MAC_string ~ MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated PMAX (W/kg)") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

The following figure depicts estimated time correction, and as can be seen, only the time and distance correction model estimated the time correction correctly (i.e., the *stolen time*; indicated by the dashed line on the figure).

>>>>

---
title: "shorts: An R Package for Modeling Short Sprints"
preprint: true
date: "`r Sys.Date()`"
author: 
  - name: Mladen Jovanović
    affiliation: 1
    corresponding: true
    email: coach.mladen.jovanovic@gmail.com
  - name: Jason D. Vescovi
    affiliation: 2
    corresponding: false
    email: vescovij@gmail.com
affiliation:
  - code: 1
    address: Faculty of Sport and Physical Education, University of Belgrade, Serbia
  - code: 2
    address: Faculty of Kinesiology and Physical Education, Graduate School of Exercise Science, Toronto, ON Canada
abstract: >
  Short sprint performance is one of the most distinguishable and admired physical traits in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). The most common methods to assess short sprint performance are with a radar gun or timing gates. In this paper, we: 1) provide the **shorts** package that can model sprint timing data from these two sources; 2) discuss potential issues with assessing sprint time (synchronization and flying start, respectively); and 3) provide model definitions within the **shorts** package to help alleviate errors within the subsequent parameter outcomes.
header-includes: >
  \usepackage{lipsum}
  \usepackage[htt]{hyphenat}
  \usepackage{booktabs}
  \usepackage{microtype}
preamble: >
  \usepackage{amsmath}
citation_package: natbib
bibliography: [references.bib, packages.bib]
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
    extra_dependencies: ["flafter"]
  rticles::peerj_article: default
link-citations: yes
editor_options:
  chunk_output_type: console
---

@@

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  fig.retina = 0.8, # figures are either vectors or 300 dpi diagrams
  dpi = 300,
  out.width = "90%",
  fig.align = "center",
  fig.width = 5,
  fig.height = 5 * 0.618, # 1 / phi
  fig.show = "hold",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  width = 65,
  size = "small"
)

# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), "bookdown", "knitr", "rmarkdown", "shorts", "nlme", "LambertW", "kableExtra"
 ), "packages.bib")

@@

Although TAU is used in the equations, and later estimated, we prefer to use MAC instead since it is easier to grasp, particularly for less math inclined coaches.

@@

Plotting acceleration against velocity (Figure \@ref(fig:four-athletes-profile)), we will get *Acceleration-Velocity Profile*, which is linear, according to the mathematical model. If the athlete's body mass (kg) is known, as well as additional air resistance parameters (see [Air resistance and the calculation of force and mechanical power] section of this paper), *Force-Velocity Profile* can be estimated (see [Force-Velocity profile] section of this paper).

```{r four-athletes-profile, echo=FALSE, fig.cap="Acceleration-Velocity profile of four athletes with different MSS and MAC parameters."}
gg <- ggplot(
  kinematics,
  aes(x = velocity, y = acceleration, color = Athlete)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  ylab(expression("Acceleration (ms"^-2 * ")")) +
  xlab(expression("Velocity (ms"^-1 * ")")) +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40 m), (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \@ref(eq:velocity-time) is performed in **shorts** package using non-linear least squares regression implemented in the `nls()` function [@batesNonlinearModels1992; @batesNonlinearRegressionAnalysis2007] in the **base R** [@R-base] and `nlme()` function in the **nlme** package [@R-nlme] for the mixed-effect models.

## Estimating short sprint parameters using timing gates split times

Let's consider an example of an athlete with MSS equal to 9 $ms^{-1}$, TAU equal to 1.3, and MAC equal to 6.92 $ms^{-2}$ performing 40m sprint with timing gates positioned at each 10m split. For split times, distance is a predictor, and time is the outcome variable, thus the equation \@ref(eq:distance-time) becomes:

@@

$W$ in equation \@ref(eq:time-distance) represents Lambert's W function [@R-LambertW]. Researchers often incorrectly use the equation \@ref(eq:distance-time) [@morinSpreadsheetSprintAcceleration2017; @morinSpreadsheetSprintAcceleration2019; @stenrothSpreadsheetSprintAcceleration2020], in which the time is the predictor and distance is the outcome variable, instead of statistically correct equation \@ref(eq:time-distance) [@motulskyIntuitiveBiostatisticsNonmathematical2018, p. 341].

MSS and TAU parameters are estimated using `model_using_splits()` function:

@@

Additional information about mixed-effects model performed using the `nlme` package [@R-nlme] can be obtained using `summary()` function. S3 method `coef()` when applied on mixed-model result will return both the fixed and random effects.

@@

To create a simple plot of the model, use S3 `plot()` method.

@@

## Problems at the start when using timing gates

@@

John represents the *theoretical model*, in which we assume that the initial force production and the timing initiation are perfectly synchronized. Jack, on the other hand, represents a *practical model*, and decides to move slightly behind the initial timing gate (i.e. for 0.5 m) and use body rocking to initiate the sprint start. In other words, Jack is using a *flying start*, a common scenario when testing field sports athletes. Flying start distance is often recommended to avoid premature triggering of the timing system by lifted knees or swinging
arms [@altmannAccuracySingleBeam2018; @altmannDifferentStartingDistances2015; @altmannValiditySingleBeamTiming2017; @haugenPowerForceVelocityProfilingSprinting2020; @haugenSprintRunningPerformance2016].

@@

Here is a graphical representation of the sprint splits (please refer to the [Supplemental material] for the R code):

@@

To explore this further, we have run a simple simulation by increasing Jack's flying start distance from 0 to 1 m and depicting the estimated MSS, TAU, MAC, and PMAX parameters (please refer to the [Supplemental material] for the R code).

```{r include=FALSE}
sim_df <- expand.grid(
  MSS = 9,
  MAC = 9 / 1.3,
  flying_start_distance = seq(0.0001, 1, length.out = 1000),
  distance = c(5, 10, 20, 30, 40)
)

@@

# Prediction residuals
model_df$residuals <-  model_df$time - model_df$predicted_time
```

```{r echo=FALSE}
# Estimates plot
plot_df <- model_df %>%
  group_by(MSS, TAU, flying_start_distance) %>%
  slice(1) %>%
  mutate(
    `MSS (m/s)` = est_MSS,
    `TAU` = est_TAU,
    `MAC (m/s/s)` = est_MAC,
    `PMAX (W)` = est_PMAX
  ) %>%
  pivot_longer(cols = c("MSS (m/s)", "TAU", "MAC (m/s/s)", "PMAX (W)")) %>%
  mutate(
    name = factor(name, levels = c("MSS (m/s)", "TAU", "MAC (m/s/s)", "PMAX (W)"))
  )

ggplot(plot_df, aes(x = flying_start_distance, y = value)) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~name, scales = "free_y") + 
  xlab("Flying start distance (m)") +
  ylab("Estimated parameter")
```

As can be seen from the figure, MSS is underestimated as flying start distance increases. MAC (and also TAU) are highly affected by the flying start distance, and from the figure we can notice that MAC is overestimated as flying start distance increases. Estimated PMAX is also overestimated as flying start distance increases.

Model residuals are also affected by the flying start distance. The shape of residuals distribution depends on number and splits utilized (e.g., 10, 20, 30, 40 m versus 5, 15, 30 m), but here we can see the effect of the flying start distance on the model residuals per split distance utilized in our simple simulation:

```{r echo=FALSE}
# Residuals
plot_df <- model_df %>%
  mutate(
    distance_string = paste0(distance, "m"),
    distance_string = factor(
      distance_string,
      levels = c("5m", "10m", "20m", "30m", "40m")
    )
  )

ggplot(
  plot_df,
  aes(
    y = residuals,
    x = flying_start_distance,
  )
) +
  theme_bw(8) +
  geom_line() +
  facet_wrap(~distance_string) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Flying start distance (m)") +
  ylab("Observed time - predicted time (s)")
```

Another way to visualize the effect of the flying start distance on split distance residuals can be found on the next figure:

```{r echo=FALSE}
# Residuals
ggplot(
  model_df,
  aes(
    y = residuals,
    x = distance,
    color = flying_start_distance,
    group = flying_start_distance
  )
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Observed - predicted time (s)") +
  theme(legend.position = "top") +
  labs(color = "Flying start distance")
```

@@

In Jack's case, both +0.3 s fixed time correction and time correction estimation yield parameters closer to John's (i.e. true parameters). We have used these two model definitions in retrospective pilot study [@vescoviSprintMechanicalCharacteristics2021], demonstrating statistically significant differences in estimated FVP parameters.

Instead of using time correction as simple intercept in the equation \@ref(eq:time-correction), we can estimate the *distance correction* (i.e., the flying start distance) using the following equation:

\begin{equation}
  \begin{split}
   t(d) &= (TAU \times W(-e^{\frac{-d + distance \; correction}{MSS \times TAU}} - 1) + \frac{d + distance \; correction}{MSS} + TAU) \\ 
   &\quad-(TAU \times W(-e^{\frac{distance \; correction}{MSS \times TAU}} - 1) + \frac{distance \; correction}{MSS} + TAU) 
   \end{split}
   (\#eq:distance-correction)
\end{equation}

Equation \@ref(eq:distance-correction) was used to generate the data for John and Jack (please refer to the provided R code), as well as for the simple simulation, and can be used as model definition to control for the flying start distance.

To estimate distance correction parameter, we use `model_using_splits_with_distance_correction()` function. Here is how we can estimate Jack parameters using distance correction:

```{r}
jack_profile_distance_correction <- model_using_splits_with_distance_correction(
  distance = split_times$distance,
  time = split_times$jack_time
)

jack_parameters <- rbind(
  jack_parameters,
  coef(jack_profile_distance_correction)
)

rownames(jack_parameters) [6] <- "Jack - Estimated distance correction"

Another model definition, which also represents a novel approach implemented in the **shorts** package, is to utilize both time and distance corrections. Thus, equation \@ref(eq:time-correction) becomes:

\begin{equation}
  t(d) = TAU \times W(-e^{\frac{-d + distance \; correction}{MSS \times TAU}} - 1) + \frac{d + distance \; correction}{MSS} + TAU - time \; correction (\#eq:time-distance-correction)
\end{equation}

@@

```{r}
jack_profile_time_distance_correction <- model_using_splits_with_corrections(
  distance = split_times$distance,
  time = split_times$jack_time
)

jack_parameters <- rbind(
  jack_parameters,
  coef(jack_profile_time_distance_correction)
)

rownames(jack_parameters) [7] <- 
  "Jack - Estimated time & distance corrections"

@@

As can be seen from the results, adding distance correction results in correctly estimating Jack's sprint parameters. There are a few issues with this model definition. Besides being novel and still not validated with actual data, distance correction model has four parameters to estimate, which implies that at least four sprint splits are needed. This imposes practical limitations, since acquiring five timing gates (one for the start and four for splits) might be practically troublesome. One strategy that is sometimes implemented is adding zeros to the sample (i.e., $t=0$ and $d=0$), which increase the number of observations. Unfortunately, this strategy should not be implemented since it will bias the estimated parameters:

```{r}
jack_profile_zero <- model_using_splits(
  distance = c(0, split_times$distance),
  time = c(0, split_times$jack_time)
)

jack_profile_time_estimated_zero <- model_using_splits_with_time_correction(
  distance = c(0, split_times$distance),
  time = c(0, split_times$jack_time)
)

adding_zero_issue <- rbind(
  coef(jack_profile),
  coef(jack_profile_zero),
  coef(jack_profile_time_estimated),
  coef(jack_profile_time_estimated_zero)
)

rownames(adding_zero_issue) <- c(
  "Jack - No corrections",
  "Jack - No corrections with zeros added",
  "Jack - Estimated time correction",
  "Jack profile - Estimated time correction with added zeros")

kable(adding_zero_issue, digits = 2, booktabs = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"))
```

Adding zeros to the sample nullifies the potential benefits of using time correction model and should be avoided in practice.

It is important to note that using `predict_XXX()` family of functions with distance correction model and using estimated distance correction parameter will yield wrong predictions on the original scale (but not on the *theoretical* scale in which force production starts at d=0m and t=0s). This is because `predict_XXX()` family of functions utilize equation \@ref(eq:time-distance-correction) and its derivates and integrals, which works for the uncorrected model (time and distance correction parameters are equal to zero), time correction models (distance correction parameter is equal to zero), and time and distance corrections model. You can notice that equations \@ref(eq:time-correction) and \@ref(eq:time-distance-correction) are equivalent to the \@ref(eq:time-distance) equation when time and distance correction parameters are equal to zero. Unfortunately, this doesn't work for the distance correction model when estimated distance correction parameter is used since model definition outlined in the equation \@ref(eq:distance-correction) is different that the equation \@ref(eq:time-distance-correction) when estimated distance correction parameter is different than zero. The solution is to use generic S3 `predict()` function, but this only predicts split times at split distances.

```{r}
# Jack's observed times
split_times$jack_time

# To predict at "theoretical" distance scale (where force application starts at d=0m)
predict_time_at_distance(
  distance = split_times$distance, 
  MSS = jack_profile_time_estimated$parameters$MSS,
  TAU = jack_profile_time_estimated$parameters$TAU)

predict_time_at_distance(
  distance = split_times$distance, 
  MSS = jack_profile_distance_correction$parameters$MSS,
  TAU = jack_profile_distance_correction$parameters$TAU)

# To predict at original distance scale, use all estimated parameters
predict_time_at_distance(
  distance = split_times$distance, 
  MSS = jack_profile_time_estimated$parameters$MSS,
  TAU = jack_profile_time_estimated$parameters$TAU,
  time_correction = jack_profile_time_estimated$parameters$time_correction)

# Unfortunately, this doesn't work for the distance correction model
predict_time_at_distance(
  distance = split_times$distance, 
  MSS = jack_profile_distance_correction$parameters$MSS,
  TAU = jack_profile_distance_correction$parameters$TAU,
  distance_correction = jack_profile_distance_correction$parameters$distance_correction)

# Instead use generic S3 predict() function
predict(jack_profile_distance_correction)

predict(
  jack_profile_distance_correction$model,
  newdata = data.frame(distance = split_times$distance))
```

In the next section we examine how these models (no correction, fixed time correction, estimated time correction, estimated distance correction, and estimated time and distance correction) perform using simulated data with varying flying start distance (please refer to the [Supplemental material] for the R code).

```{r echo=FALSE}
pred_wrapper <- function(data) {

  merge_model <- function(data, model, model_name) {
    data$model <- model_name
    data$est_MSS <- model$parameters$MSS
    data$est_TAU <- model$parameters$TAU
    data$est_MAC <- model$parameters$MAC
    data$est_PMAX <- model$parameters$PMAX
    data$est_time_correction <- model$parameters$time_correction
    data$est_distance_correction <- model$parameters$distance_correction
    data$predicted_time <- model$data$pred_time
    data$residuals <- data$time - data$predicted_time

    data
  }

  no_correction <- model_using_splits(
    distance = data$distance,
    time = data$time
  )

@@

  time_correction <- model_using_splits_with_time_correction(
    distance = data$distance,
    time = data$time,
    control = nls.control(warnOnly = TRUE)
  )

  distance_correction <- model_using_splits_with_distance_correction(
    distance = data$distance,
    time = data$time,
    control = nls.control(warnOnly = TRUE)
  )

  time_dist_correction <- model_using_splits_with_corrections(
    distance = data$distance,
    time = data$time,
    control = nls.control(warnOnly = TRUE)
  )

  rbind(
    merge_model(
      data, 
      no_correction,
      "No correction"
    ),
    merge_model(
      data, 
      fixed_correction_short,
      "Fixed correction +0.3s"
    ),
    merge_model(
      data, 
      fixed_correction_long,
      "Fixed correction +0.5s"
    ),
    merge_model(
      data, 
      time_correction,
      "Time correction"
    ),
    merge_model(
      data, 
      distance_correction,
      "Distance correction"
    ),
    merge_model(
      data, 
      time_dist_correction,
      "Time and distance correction"
    )
  )
}

# estimated parameters and predicted time
model_df <- sim_df %>%
  filter(flying_start_distance > 0) %>%
  group_by(MSS, TAU, flying_start_distance) %>%
  do(pred_wrapper(.)) %>%
  ungroup()
```

As can be seen from the next figure, the estimated time correction model estimates MSS almost perfectly, while the estimated distance and estimate time and distance correction models estimates MSS perfectly.

```{r echo=FALSE}
model_df$model <- factor(
  model_df$model,
  levels = c(
    "No correction",
    "Fixed correction +0.3s",
    "Fixed correction +0.5s",
    "Time correction",
    "Distance correction",
    "Time and distance correction"
  )
)

# Estimates plot
plot_df <- model_df %>%
  group_by(model, MSS, TAU, flying_start_distance) %>%
  slice(1) %>%
  mutate(
    `MSS (m/s)` = est_MSS,
    `TAU` = est_TAU,
    `MAC (m/s/s)` = est_MAC,
    `PMAX (W)` = est_PMAX
  ) %>%
  pivot_longer(cols = c("MSS (m/s)", "TAU", "MAC (m/s/s)", "PMAX (W)")) %>%
  mutate(
    name = factor(name, levels = c("MSS (m/s)", "TAU", "MAC (m/s/s)", "PMAX (W)"))
  )

ggplot(plot_df, aes(x = flying_start_distance, y = value, group = model, color = model)) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~name, scales = "free_y") + 
  xlab("Flying start distance (m)") +
  ylab("Estimated parameter") +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

Similar outcomes are observed for the MAC parameter. The distance correction model and time and distance corrections models performs perfectly, while the time correction model performs almost as good. PMAX demonstrates the same properties as MSS and MAC.

The next figure depicts residuals across split distances for each simulated flying start distance. As can be seen, time correction, distance correction, and time and distance correction models perform much better than the other models.

```{r}
# Residuals
ggplot(
  model_df,
  aes(
    y = residuals,
    x = distance,
    color = flying_start_distance,
    group = flying_start_distance
  )
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  facet_wrap(~model) +
  xlab("Distance (m)") +
  ylab("Observed - predicted time (s)") +
  theme(legend.position = "top") +
  labs(color = "Flying start distance")
```

The outcomes from this simple simulation data clearly demonstrates that the time correction, distance correction, and time and distance correction models represent sound improvements in parameter estimation and model fit compared to no corrections model and fixed correction model when attempting to overcome the flying start issues. Since the time correction and distance correction models are simpler and requires three parameters to be estimated, they might be practically more useful than the time and distance correction model, which requires four parameters estimation and thus more than four timing gates and sprint splits. More elaborate simulation study is currently ongoing and the results will be reported in another paper.
